\section{Planned Intervention}

Students have a hard time understanding existing datasets, the tools to connect to them

With the development of the new technology, interventions will be staged through Virginia Tech's new ``Introduction to Computational Thinking'' course, created to help fulfill the university's new General Education Requirements~\cite{vt-vision}. 
This course has already been run for two semesters, deeply incorporating much of my existing research work.
The course is taught and developed primarily by Dr. Dennis Kafura, although I have also been involved as associate instructor, managing course materials, server and technology administration, and assisting with in-class teaching.
However, now that the course is more solidly defined, my role is shifting into a more observational function in order to drive my dissertation work.
As a research endeavor, the course is heavily instrumented to provide data on its novel pedagogies and technologies.
Although there are confounding factors to working with such a heavily experimental course, it presents a unique testbed for materials and is an excellent source for mining research results.

\subsection{The Learners}

Introductory Computing has a number of target audiences, including the undergraduate level and the K-12 level.
Many undergraduate non-majors taking Computational Thinking, a primary target of my research efforts.
These students present a unique profile.
A few of them will have had prior programming experience, but most of them have had very minimal interactions with computers (indeed, some will describe themselves as ``not a computer person'').
These students may not believe that Computational Thinking will help them.
This is largely because they have more clearly defined domain identities, and may not see how Computational Thinking fits into them.
So, indeed, these students often have low motivation and low ability.

In the first offering of the course, 25 students enrolled in the course, and 20 students finished the coursework.
In the second offering, 40 students initially enrolled and 35 successfully passed the course.
Figure \ref{data-demographics} indicates the relevant demographic data collected through surveys.
Largely, the students represent the population at Virginia Tech, albeit with some disproportion towards certain majors than others.
It is worth pointing out the excellent gender diversity within the class.

\begin{figure*}
\begin{minipage}{\linewidth}
\centering
	\begin{tabular}{c|c|c}
	  \multicolumn{3}{c}{Gender}\\\hline
		& Fall 2014 & Spring 2015 \\\hline
		Female & 6 & 21 \\
		Male & 13 & 18 \\
	\end{tabular}
\centering
	\begin{tabular}{c|c|c}
	  \multicolumn{3}{c}{Prior Programming Experience}\\\hline
		& Fall 2014 & Spring 2015 \\\hline
		Yes & 10 & 7 \\
		No & 10 & 32 \\
	\end{tabular}
\centering
	\begin{tabular}{c|c|c}
		\multicolumn{3}{c}{Year}\\\hline
		& Fall 2014 & Spring 2015 \\\hline
		Freshman & 2 & 5 \\
		Sophomore & 7 & 11 \\
		Junior & 6 & 11 \\
		Senior & 5 & 10 \\
		Unknown & 0 & 2 \\
	\end{tabular}
\centering	
	\begin{tabular}{c|c|c}
	  \multicolumn{3}{c}{Colleges}\\\hline
		& Fall 2014 & Spring 2015 \\\hline
		Engineering & 2 & 2 \\
		Agriculture & 0 & 1 \\
		Sciences & 7 & 5 \\
		Liberal Arts & 9 & 23 \\
		Architecture & 1 & 7 \\
		Natural Resources & 0 & 0 \\
	\end{tabular}
\caption{Demographic Data of Computational Thinking Students}
\label{data-demographics}
\end{minipage}
\end{figure*}


\subsection{The Content}

Virginia Tech defines four learning objectives for computational thinking. Students are required to:
\begin{enumerate}
	\item Apply computational methods to model and analyze complex or large scale phenomena.
	\item Formulate problems and find solutions using computational or quantitative thinking in their field of study.
	\item Give examples of the application to, and discuss the significance of, computational thinking in at least two different knowledge domains. 
	\item Evaluate the social and political impact of computing and information technologies 
\end{enumerate}

This content is mapped roughly into four instructional units on Computational Modelling, Algorithms, Data Intensive Inquiry, and Social Impacts. The lattermost unit is threaded throughout the course, while the first three are roughly sequential. Figure \ref{course-outline} gives a high-level overview of the content of this course.

\begin{figure*}
\begin{tabularx}{\textwidth}{ |l|X| }
\hline
Topic (Length) &	Description \\\hline
Computational Modeling \newline\newline
 (2 weeks) & Model-based investigation of how complex global behavior arises from the interaction of many “agents”, each operating according to local rules. Students use case-based reasoning and encounter basic computation constructs in a highly supportive simulation environment. \\\hline
Fundamentals of Algorithms \newline (4 weeks) & Study of the basic constructs of programming logic (sequence, decisions, and iteration) and program organization (procedures). A block-based programming language is used to avoid syntactic details. Students can see how these constructs are expressed in Python. \\\hline
Data-intensive Inquiry \newline (7 weeks) & Project-based exploration of complex phenomena by algorithmically manipulating large-scale data from real-world sources. Students construct algorithms in Python using a supportive framework for accessing the data. \\\hline
Social Impacts \newline (2 weeks) & Explore and discuss contemporary societal issues involving computing and information technology. \\\hline
\end{tabularx}
\caption{High-Level Course Overview}
\label{course-outline}
\end{figure*}


\subsection{The Course}

The course uses a considerable amount of modern pedagogical techniques, many of which represent ongoing research questions.
Perhaps the most influential technique is the organization of students into cohorts.
Near the beginning of the semester, students are put into groups of 5-6, balancing based on year and gender where possible, and avoiding putting similar majors into the same group.
These cohorts primarily function as a support structure that students can rely on to get help and encouragement.
Although cohorts work together on many smaller in-class assignments, every student is ultimately responsible for their own work -- the final project, for instance, is individual to each student.

Class time is split between presentation (typically stand-up lecture) and participation (typically computer-based work or cohort discussion) using an Active Learning style whereever possible.
Earlier questions in the course often have students completing questions on paper or doing more kinetic exercises.
Later questions rely on the automated Kennel questions, until the students reach the open-ended project work.

Work in the class is considered to be under a mastery style -- deadlines are enforced only in so far as they motivate students to complete assignments, not in order to punish.
Students are free to work on the material as long as they need.
A recurring message within the course is that ``failing is okay, as long as you keep trying''.

\subsection{The Research Protocol}

The Computational Thinking course is deeply integrated into my research technology -- the middle section of the course uses Kennel as the programming environment, and the last section of the course uses CORGIS to manage the students' projects.
Every semester, students will be surveyed at the start, middle, and end of the course in order to collect relevant data.
Their interactions with the technological systems will be tracked at a fine-grained level, as will be their final projects.
This data can be compared across semesters as it accumulates in order to hit instructor-set goals.
Along the way, important questions related to secondary research goals can also be answered -- how to design CORGIS libraries, the affordances of Data Science as an educational tool, and the pedagogical strategies for using Kennel.

\subsection{Measuring Success}

There are three primary dimensions used in measuring the success of my proposed work: students' motivation, students' ability, and the courses' scalability.
Different instruments and sources will be used to track and measure the impact of the various research efforts.
Ultimately, gains are expected as materials are refined, ideally achieving some established performance objectives.

\subsubsection{Motivation}
Motivation in the course will be primarily assessed through self-reported surveys based heavily on the MUSIC Model of Academic Motivation Inventory (MMAMI).
Students will be non-anonymously surveyed using the complete instrument at key points in the semester, and regularly surveyed using a miniaturized version throughout the semester.
These miniaturized versions will be used as follow-ups to specific course exercises in order to determine more finely-grained answers about where students found motivation.
Additionally, targeted qualitative data will be gathered in order to refine this quantitative data.

This motivational data will be combined with evidence of engagement with the course across several different dimensions. Therefore, the complete list of motivational sources is:
\begin{description}
	\item[Attendance:] Did the students regularly attend class?
	\item[Retention:] Did the students complete the course with a passing grade?
	\item[Coursework:] Did the students regularly complete coursework and homework on time?
	\item[Struggle:] Did the students continue with materials even after setbacks?
	\item[Continuation:] Do students intend to follow-up the course with further courses and learning experiences in Computational Thinking?
	\item[Observations:] Did the course staff report students as being motivated, according to their own observations?
	\item[Self-report:] Did the students report themselves as being motivated, according to MMAMI and other surveys?
\end{description}
